Likelihoods for a general class of ARGs under the SMC

### 2025-04-29 paper.tex
Cover tricky tsinfer points

Fig.~\ref{fig:lik-surface} also shows results for ARGs inferred from 
the  \revpoint{2}{35}
simulated data.
To do this we first simulated mutations on the full ARG 
at rate $1.25\times10^{-8}$ using msprime, resulting in an 
ARG with 2,989 mutations.
We then used tsinfer and tsdate~\citep{wohns_unified_2022}
to infer a dated ARG,
using default parameters for tsinfer and the true values of mutation
rate and $N_e$ for tsdate.
In the first, we use the default output of tsinfer
(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations). \revpoint{2}{36}
This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
with three or more children per tree. Tsinfer does not directly estimate
recombination events; the breakpoints between trees are the consequence of
switches between parental haplotypes in the
\cite{li2003modeling} copying process. Information about
recombination events in the returned ARG is therefore quite imprecise,
and likely to contain incorrect and contradictory details if one were to
attempt to reconstruct the exact subtree-prune-and-regraft
moves~\citep[e.g.,][]{rasmussen2023espalier}. Nonetheless, in
this toy example at least, our likelihood function performs well \revpoint{2}{37}
on the output of tsinfer, with the minimum close to the true
parameter value. We also show the likelihood curve obtained when we
fully simplify the output of tsinfer
(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
before dating. We can see that, as with the true data,
removing all the unary nodes results in a bias \revpoint{2}{38}
away from the true value of the recombination rate parameter.

>>>>

Fig.~\ref{fig:lik-surface} also shows results for ARGs inferred from 
the  \revpoint{2}{35}
simulated data.
To do this we first simulated mutations on the full ARG 
at rate $1.25\times10^{-8}$ using msprime, resulting in an 
ARG with 2,989 mutations.
We then used tsinfer and tsdate~\citep{wohns_unified_2022}
to infer a dated ARG,
using default parameters for tsinfer and the true values of mutation
rate and $N_e$ for tsdate.
In the first example, we use the default output of tsinfer
(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations). \revpoint{2}{36}
This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
with three or more children per tree. Tsinfer does not directly estimate
recombination events; the breakpoints between trees are the consequence of
switches between parental haplotypes in the
\cite{li2003modeling} copying process. Information about
recombination events in the returned ARG is therefore quite imprecise,
and likely to contain incorrect and contradictory details if one were to
attempt to reconstruct the exact subtree-prune-and-regraft
moves~\citep[e.g.,][]{rasmussen2023espalier}. It is encouraging,
therefore, that in this toy example our likelihood function is robust
to these omitted and incorrect details in \revpoint{2}{37}
the output of tsinfer, with the minimum close to the true
parameter value. We also show the likelihood curve obtained when we
fully simplify the output of tsinfer
(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
before dating. We can see that, as with the true data,
removing all the unary nodes results in a bias \revpoint{2}{38}
away from the true value of the recombination rate parameter.
